---
title: "Ordinary least squares regression"
author: "Ruixuan Tu (ruixuan.tu@wisc.edu)"
date: "13 February 2023"
---

# Introduction

Implementing ordinary least squares regression from scratch for 1D case. This homework is assigned on 8 February 2023.

- Materials
  - https://en.wikipedia.org/wiki/Ordinary_least_squares
  - https://www.cs.princeton.edu/courses/archive/fall18/cos324/files/linear-regression.pdf
  - https://setosa.io/ev/ordinary-least-squares-regression/
- Dataset
  - https://www.kaggle.com/datasets/siddheshmahajan/world-gdp-over-the-years

# Setup

```{r results='hide', message=FALSE, warning=FALSE}
knitr::opts_knit$set(root.dir = "/Users/turx/Projects/machine-teaching-23sp/hw01-ols")
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(readxl)
library(tidyverse)
library(lubridate)
library(ggplot2)
data <- read_excel("Data_Extract_From_World_Development_Indicators.xlsx")
data <- data %>%
    na_if("..") %>%
    rename(c_name = "Country Name", c_code = "Country Code", "2017" = "2017 [YR2017]", "2018" = "2018 [YR2018]") %>%
    pivot_longer(cols = paste(c(1990, 2000, c(2010:2019))), names_to = "year", values_to = "gdp") %>%
    filter(`Series Name` == "GDP (current US$)", `Series Code` == "NY.GDP.MKTP.CD") %>%
    select(-`Series Name`, -`Series Code`) %>%
    drop_na() %>%
    mutate(gdp = as.numeric(gdp), year = as.numeric(year)) %>%
    group_by(c_code)
```

# Definition of OLS Regression Function on 1D

- input: vectors $\vec{x}, \vec{y}$
- output: a list with two elements: $m$ and $b$, the slope and intercept of the regression line $y=mx+b$

```{r}
ols_regression <- function(x, y) {
    x <- as.matrix(x)
    y <- as.matrix(y)
    m <- (mean(x * y) - mean(x) * mean(y)) / (mean(x^2) - mean(x)^2)
    b <- mean(y) - m * mean(x)
    return(list(m = m, b = b))
}
```

# Demonstration of OLS Regression on USA GDP

```{r}
usa <- data %>%
    filter(c_code == "USA") %>%
    ungroup() %>%
    select(year, gdp)

usa %>%
    ggplot(aes(x = year, y = gdp)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = "GDP of the United States (with library)", x = "Year", y = "GDP", color = "blue") +
    scale_y_continuous(labels = scales::dollar)

l_ols <- ols_regression(usa$year, usa$gdp)
usa %>%
    ggplot(aes(x = year, y = gdp)) +
    geom_point() +
    geom_abline(intercept = l_ols$b, slope = l_ols$m, color = "red") +
    labs(title = "GDP of the United States (without library)", x = "Year", y = "GDP") +
    scale_y_continuous(labels = scales::dollar)
```

The learned coefficients for $y=mx+b$

```{r}
l_ols
```

Training set error: $E_{\text{train}} = \sum_{i=1}^n (y_i - \hat{y}_i)^2$

```{r}
error <- usa %>%
    mutate(y_hat = l_ols$m * year + l_ols$b) %>%
    mutate(error = (gdp - y_hat)^2) %>%
    summarise(error = sum(error)) %>%
    pull()
error
```

MSE: $MSE = \frac{1}{n} E_{\text{train}}$

```{r}
MSE <- error / nrow(usa)
MSE
```
